---
title: "Exercise 5: Senior Debt Share in LICs"
subtitle: "BP3 — Fund's Role in International Debt Architecture"
output: html_notebook
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(openxlsx)
library(scales)
cat("✓ Libraries loaded\n")
```

```{r paths}
ids_dir             <- "U:/dil/input/IDS"
characteristics_dir <- "U:/dil/main-characteristics"
output_dir          <- "U:/dil/bp3/output"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# Load and combine all 6 IDS files

```{r load-ids}
files <- c("A_D.xlsx", "E_K.xlsx", "L_M.xlsx", "N_Q.xlsx", "R_U.xlsx", "V_Z.xlsx")

ids_list <- list()
for (f in files) {
  cat("Loading", f, "...\n")
  ids_list[[f]] <- read.xlsx(file.path(ids_dir, f))
  cat("  ✓", format(nrow(ids_list[[f]]), big.mark = ","), "rows\n")
}

year_cols <- names(ids_list[[1]])[grepl("^\\d{4}$", names(ids_list[[1]]))]

for (i in seq_along(ids_list)) {
  ids_list[[i]][year_cols] <- lapply(ids_list[[i]][year_cols], as.character)
}

ids_raw <- bind_rows(ids_list)
rm(ids_list)
gc()

cat("✓ Combined:", format(nrow(ids_raw), big.mark = ","), "rows\n")
cat("Countries:", n_distinct(ids_raw$Country.Name), "\n")
cat("Series:", n_distinct(ids_raw$Series.Code), "\n")
```

---

# Filter to key series only

We need three series:

- `DT.DOD.MLAT.CD` — PPG multilateral DOD (to extract PCS subset by counterpart area)
- `DT.DOD.DIMF.CD` — IMF credit DOD (use of Fund credit)
- `DT.DOD.DPPG.CD` — Total PPG DOD (denominator base)

```{r filter-series}
keep_series <- c(
  "DT.DOD.MLAT.CD",   # PPG multilateral DOD
  "DT.DOD.DIMF.CD",   # IMF credit DOD
  "DT.DOD.DPPG.CD"    # Total PPG DOD
)

ids_raw <- ids_raw %>% filter(Series.Code %in% keep_series)
cat("Filtered to key series:", format(nrow(ids_raw), big.mark = ","), "rows\n")
```

---

# Remove regional aggregates

```{r remove-aggregates}
remove_these <- c(
  "East Asia & Pacific (excluding high income)",
  "Europe & Central Asia (excluding high income)",
  "IDA only",
  "IDA total",
  "Latin America & Caribbean (excluding high income)",
  "Least developed countries: UN classification",
  "Low & middle income",
  "Low income",
  "Lower middle income",
  "Middle East & North Africa (excluding high income)",
  "Middle East, North Africa, Afghanistan & Pakistan (excluding high income)",
  "Middle income",
  "South Asia",
  "Sub-Saharan Africa (excluding high income)",
  "Upper middle income",
  "Heavily indebted poor countries (HIPC)",
  "World"
)

ids_raw <- ids_raw %>% filter(!Country.Name %in% remove_these)
cat("After removing aggregates:", n_distinct(ids_raw$Country.Name), "countries\n")
```

---

# Reshape to long format

```{r reshape}
year_cols <- names(ids_raw)[grepl("^\\d{4}$", names(ids_raw))]

ids_long <- ids_raw %>%
  pivot_longer(cols = all_of(year_cols), names_to = "year", values_to = "value") %>%
  mutate(year  = as.integer(year),
         value = as.numeric(value)) %>%
  filter(year >= 1970, year <= 2024, !is.na(value)) %>%
  rename(country_code = Country.Code,
         country_name = Country.Name,
         ca_name      = `Counterpart-Area.Name`,
         ca_code      = `Counterpart-Area.Code`,
         series_name  = Series.Name,
         series_code  = Series.Code)

ids_long <- ids_long %>% mutate(ca_name = trimws(ca_name))
ids_long <- ids_long %>% mutate(country_code = trimws(country_code))


rm(ids_raw)
gc()
cat("✓ ids_long:", format(nrow(ids_long), big.mark = ","), "rows\n")
```

---

# LIC classification (country-characteristics.xlsx)

```{r lic-classification}
chars   <- read.xlsx(file.path(characteristics_dir, "country-characteristics.xlsx"), sheet = "main")
lic_iso <- trimws(chars$iso)
lic_iso <- lic_iso[!is.na(lic_iso) & lic_iso != ""]

ids_long <- ids_long %>%
  mutate(income_group = if_else(country_code %in% lic_iso, "LIC", "MIC"))

cat("LIC:", n_distinct(ids_long$country_code[ids_long$income_group == "LIC"]),
    "| MIC:", n_distinct(ids_long$country_code[ids_long$income_group == "MIC"]), "\n")

```

---

# Define PCS institutions

```{r pcs-codes}
pcs_codes <- c(
  "901",  # World Bank-IBRD
  "905",  # World Bank-IDA
  "902",  # World Bank-MIGA
  "903",  # IFC
  "988",  # IFAD
  "911",  # BIS
  "990",  # EBRD
  "913",  # African Dev. Bank
  "915",  # Asian Dev. Bank
  "909",  # Inter-American Dev. Bank
  "899",  # AIIB
  "926",  # Council of Europe
  "906",  # Caribbean Dev. Bank
  "919",  # European Investment Bank
  "957",  # BOAD
  "887",  # NDB
  "976",  # Islamic Dev. Bank
  "920",  # CAF
  "951",  # OPEC Fund
  "973",  # Arab Monetary Fund
  "953"   # BADEA
)

# Verify PCS codes match multilateral counterpart areas
mlat_ca <- ids_long %>% filter(series_code == "DT.DOD.MLAT.CD") %>% distinct(ca_code)
cat("PCS matched:", sum(pcs_codes %in% mlat_ca$ca_code), "of", length(pcs_codes), "\n")
unmatched <- pcs_codes[!pcs_codes %in% mlat_ca$ca_code]
if (length(unmatched) > 0) cat("Unmatched PCS codes:", paste(unmatched, collapse = ", "), "\n")
```

---

# Build components

## PCS multilateral debt

```{r pcs-multilateral}
# Sum individual PCS institutions per country-year
multilateral_pcs <- ids_long %>%
  filter(series_code == "DT.DOD.MLAT.CD",
         ca_code %in% pcs_codes) %>%
  group_by(country_name, country_code, year, income_group) %>%
  summarise(pcs_value = sum(value, na.rm = TRUE), .groups = "drop")

cat("PCS multilateral rows:", nrow(multilateral_pcs), "\n")
```

## IMF credit

```{r imf-credit}
imf <- ids_long %>%
  filter(series_code == "DT.DOD.DIMF.CD", ca_code == "WLD") %>%
  select(country_name, country_code, year, income_group, imf_value = value)

cat("IMF credit rows:", nrow(imf), "\n")
```

## Total PPG

```{r total-ppg}
total_ppg <- ids_long %>%
  filter(series_code == "DT.DOD.DPPG.CD", ca_code == "WLD") %>%
  select(country_name, country_code, year, income_group, total_ppg_value = value)

cat("Total PPG rows:", nrow(total_ppg), "\n")
```

---

# Merge and compute senior debt share

Senior debt = PCS multilateral + IMF credit

Denominator = Total PPG + IMF credit (since IMF credit is separate from PPG in IDS methodology)

```{r senior-share}
senior <- total_ppg %>%
  left_join(multilateral_pcs,
            by = c("country_name", "country_code", "year", "income_group")) %>%
  left_join(imf,
            by = c("country_name", "country_code", "year", "income_group")) %>%
  mutate(
    pcs_value       = replace_na(pcs_value, 0),
    imf_value       = replace_na(imf_value, 0),
    total_ppg_value = replace_na(total_ppg_value, 0),
    senior_debt     = pcs_value + imf_value,
    denominator     = total_ppg_value + imf_value
  ) %>%
  filter(denominator > 0) %>%
  mutate(senior_share = senior_debt / denominator * 100)

cat("Senior share rows:", nrow(senior), "\n")
cat("Year range:", min(senior$year), "-", max(senior$year), "\n")
```

---

# Quick diagnostics

```{r diagnostics}
# Check distribution
senior %>%
  filter(income_group == "LIC") %>%
  summarise(
    n        = n(),
    min      = min(senior_share),
    q1       = quantile(senior_share, 0.25),
    median   = median(senior_share),
    mean     = mean(senior_share),
    q3       = quantile(senior_share, 0.75),
    max      = max(senior_share)
  )

# Country-year counts per year (LICs)
senior %>%
  filter(income_group == "LIC") %>%
  count(year) %>%
  print(n = Inf)
```

---

# Aggregate: median, mean, IQR by year — LICs

```{r aggregate-lic}
senior_agg_lic <- senior %>%
  filter(income_group == "LIC") %>%
  group_by(year) %>%
  summarise(
    median_share = median(senior_share, na.rm = TRUE),
    mean_share   = mean(senior_share, na.rm = TRUE),
    q1_share     = quantile(senior_share, 0.25, na.rm = TRUE),
    q3_share     = quantile(senior_share, 0.75, na.rm = TRUE),
    n_countries  = n(),
    .groups      = "drop"
  )

head(senior_agg_lic, 10)
```

---

# Chart: Senior Debt Share in LICs

```{r chart-lic, fig.width=9, fig.height=5.5}
ggplot(senior_agg_lic, aes(x = year)) +
  geom_ribbon(aes(ymin = q1_share, ymax = q3_share),
              alpha = 0.12, fill = "grey50") +
  geom_line(aes(y = median_share, color = "Median"),  linewidth = 0.9) +
  geom_line(aes(y = mean_share,   color = "Average"), linewidth = 0.9) +
  scale_color_manual(
    values = c("Median" = "steelblue", "Average" = "orange")
  ) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  scale_y_continuous(breaks = seq(0, 100, 10)) +
  labs(
    title = "Senior Debt Share in LICs (% of Total External PPG Debt)",
    x     = "Year",
    y     = "Percent",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```


---

# Export Excel

```{r Export}
wb_out <- createWorkbook()

addWorksheet(wb_out, "Chart - LIC")
writeData(wb_out, "Chart - LIC", senior_agg_lic)

addWorksheet(wb_out, "Detail - LIC")
writeData(wb_out, "Detail - LIC",
  senior %>%
    filter(income_group == "LIC") %>%
    select(country_name, country_code, year,
           pcs_value, imf_value, total_ppg_value,
           senior_debt, denominator, senior_share) %>%
    arrange(country_name, year))

excel_file <- file.path(output_dir, "ex5_senior_debt_share.xlsx")
saveWorkbook(wb_out, excel_file, overwrite = TRUE)
cat("✓ Saved:", excel_file, "\n")
```

---

# (Optional) Aggregate for MICs and All countries

```{r aggregate-mic-all}
# MIC
senior_agg_mic <- senior %>%
  filter(income_group == "MIC") %>%
  group_by(year) %>%
  summarise(
    median_share = median(senior_share, na.rm = TRUE),
    mean_share   = mean(senior_share, na.rm = TRUE),
    q1_share     = quantile(senior_share, 0.25, na.rm = TRUE),
    q3_share     = quantile(senior_share, 0.75, na.rm = TRUE),
    n_countries  = n(),
    .groups      = "drop"
  )

# All
senior_agg_all <- senior %>%
  group_by(year) %>%
  summarise(
    median_share = median(senior_share, na.rm = TRUE),
    mean_share   = mean(senior_share, na.rm = TRUE),
    q1_share     = quantile(senior_share, 0.25, na.rm = TRUE),
    q3_share     = quantile(senior_share, 0.75, na.rm = TRUE),
    n_countries  = n(),
    .groups      = "drop"
  )
```

---

# Export to Excel

```{r export-excel}
wb_out <- createWorkbook()

# --- Chart data ---
addWorksheet(wb_out, "Chart - LIC")
writeData(wb_out, "Chart - LIC", senior_agg_lic)

addWorksheet(wb_out, "Chart - MIC")
writeData(wb_out, "Chart - MIC", senior_agg_mic)

addWorksheet(wb_out, "Chart - All")
writeData(wb_out, "Chart - All", senior_agg_all)

# --- Country-level detail (LICs only) ---
addWorksheet(wb_out, "Detail - LIC")
writeData(wb_out, "Detail - LIC",
  senior %>%
    filter(income_group == "LIC") %>%
    select(country_name, country_code, year,
           pcs_value, imf_value, total_ppg_value,
           senior_debt, denominator, senior_share) %>%
    arrange(country_name, year))

# --- Reference: PCS institutions ---
addWorksheet(wb_out, "PCS Institutions")
writeData(wb_out, "PCS Institutions",
  data.frame(ca_code = pcs_codes) %>%
    left_join(
      ids_long %>%
        filter(series_code == "DT.DOD.MLAT.CD") %>%
        distinct(ca_code, ca_name),
      by = "ca_code") %>%
    arrange(ca_name))

# --- Reference: LIC list ---
addWorksheet(wb_out, "LIC Countries")
writeData(wb_out, "LIC Countries",
  senior %>%
    filter(income_group == "LIC") %>%
    distinct(country_name, country_code) %>%
    arrange(country_name))

excel_file <- file.path(output_dir, "ex5_senior_debt_share.xlsx")
saveWorkbook(wb_out, excel_file, overwrite = TRUE)
cat("✓ Saved:", excel_file, "\n")
```

---

# Save chart

```{r save-chart}
p <- ggplot(senior_agg_lic, aes(x = year)) +
  geom_ribbon(aes(ymin = q1_share, ymax = q3_share),
              alpha = 0.12, fill = "grey50") +
  geom_line(aes(y = median_share, color = "Median"),  linewidth = 0.9) +
  geom_line(aes(y = mean_share,   color = "Average"), linewidth = 0.9) +
  scale_color_manual(
    values = c("Median" = "steelblue", "Average" = "orange")
  ) +
  scale_x_continuous(breaks = seq(1970, 2025, 5)) +
  scale_y_continuous(breaks = seq(0, 100, 10)) +
  labs(
    title = "Senior Debt Share in LICs (% of Total External PPG Debt)",
    x     = "Year",
    y     = "Percent",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

ggsave(file.path(output_dir, "ex5_senior_debt_share_lic.png"),
       p, width = 9, height = 5.5, dpi = 300)
cat("✓ Chart saved\n")
```
