---
title: "Exercise 7: Debt Resolution Mechanisms"
subtitle: "BP3 — Fund's Role in International Debt Architecture"
output: html_notebook
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(stringr)
library(openxlsx)
library(scales)
cat("Libraries loaded\n")
```

```{r paths}
ids_dir             <- "../ids"
characteristics_dir <- "."
output_dir          <- "output/ex7"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Load IDS (6 Excel files)

```{r load-ids}
files <- c("A_D.xlsx", "E_K.xlsx", "L_M.xlsx", "N_Q.xlsx", "R_U.xlsx", "V_Z.xlsx")
ids_list <- list()
for (f in files) {
  cat("Loading", f, "...\n")
  ids_list[[f]] <- read.xlsx(file.path(ids_dir, f))
  cat("  done:", format(nrow(ids_list[[f]]), big.mark = ","), "rows\n")
}
year_cols <- names(ids_list[[1]])[grepl("^\\d{4}$", names(ids_list[[1]]))]
for (i in seq_along(ids_list)) {
  ids_list[[i]][year_cols] <- lapply(ids_list[[i]][year_cols], as.character)
}
ids_raw <- bind_rows(ids_list)
rm(ids_list); gc()
cat("\nCombined:", format(nrow(ids_raw), big.mark = ","), "rows |",
    n_distinct(ids_raw$Country.Name), "countries\n")
```

---

# 2. Filter to key series

```{r filter-series}
# Private debt uses dedicated series per type (NOT counterpart filtering):
#   DT.DOD.PBND.CD = PPG bonds
#   DT.DOD.PCBK.CD = PPG commercial banks
#   DT.DOD.PROP.CD = PPG other private creditors
keep_series <- c(
  "DT.DOD.BLAT.CD",   # PPG bilateral (by counterpart — needed for PC vs non-PC split)
  "DT.DOD.MLAT.CD",   # PPG multilateral
  "DT.DOD.PBND.CD",   # PPG bonds
  "DT.DOD.PCBK.CD",   # PPG commercial banks
  "DT.DOD.PROP.CD",   # PPG other private creditors
  "DT.DOD.DPPG.CD",   # Total PPG (denominator base)
  "DT.DOD.DIMF.CD"    # IMF credit (added to denominator; always outside framework)
)

ids_raw <- ids_raw %>% filter(Series.Code %in% keep_series)
cat("Filtered:", format(nrow(ids_raw), big.mark = ","), "rows\n")
ids_raw %>% count(Series.Code, Series.Name) %>% as.data.frame() %>% print()
```

---

# 3. Remove aggregates

```{r remove-aggregates}
remove_these <- c(
  "East Asia & Pacific (excluding high income)",
  "Europe & Central Asia (excluding high income)",
  "IDA only", "IDA total",
  "Latin America & Caribbean (excluding high income)",
  "Least developed countries: UN classification",
  "Low & middle income", "Low income", "Lower middle income",
  "Middle East & North Africa (excluding high income)",
  "Middle East, North Africa, Afghanistan & Pakistan (excluding high income)",
  "Middle income", "South Asia",
  "Sub-Saharan Africa (excluding high income)",
  "Upper middle income",
  "Heavily indebted poor countries (HIPC)", "World"
)
ids_raw <- ids_raw %>% filter(!Country.Name %in% remove_these)
cat("Countries after removing aggregates:", n_distinct(ids_raw$Country.Name), "\n")
```

---

# 4. Reshape to long format

```{r reshape}
year_cols <- names(ids_raw)[grepl("^\\d{4}$", names(ids_raw))]

ids_long <- ids_raw %>%
  pivot_longer(cols = all_of(year_cols), names_to = "year", values_to = "value") %>%
  mutate(year  = as.integer(year),
         value = as.numeric(value)) %>%
  filter(year >= 1970, year <= 2024, !is.na(value)) %>%
  rename(country_code = Country.Code,
         country_name = Country.Name,
         ca_name      = `Counterpart-Area.Name`,
         ca_code      = `Counterpart-Area.Code`,
         series_name  = Series.Name,
         series_code  = Series.Code) %>%
  mutate(ca_code      = trimws(ca_code),
         country_code = trimws(country_code))

rm(ids_raw); gc()
cat("ids_long:", format(nrow(ids_long), big.mark = ","), "rows\n")
```

---

# 5. LIC classification

```{r lic-classification}
chars   <- read.xlsx(file.path(characteristics_dir, "country-characteristics.xlsx"), sheet = "main")
lic_iso <- trimws(chars$iso)
lic_iso <- lic_iso[!is.na(lic_iso) & lic_iso != ""]

ids_long <- ids_long %>%
  mutate(income_group = if_else(country_code %in% lic_iso, "LIC", "MIC"))

cat("LIC:", n_distinct(ids_long$country_code[ids_long$income_group == "LIC"]),
    "| MIC:", n_distinct(ids_long$country_code[ids_long$income_group == "MIC"]), "\n")
```

---

# 6. Paris Club mapping

```{r pc-mapping}
pc_mapping <- tribble(
  ~ca_code,  ~pc_name,              ~accession_year,
  "801",     "Australia",           1956,
  "001",     "Austria",             1956,
  "002",     "Belgium",             1956,
  "301",     "Canada",              1956,
  "003",     "Denmark",             1956,
  "018",     "Finland",             1956,
  "004",     "France",              1956,
  "005",     "Germany, Fed. Rep.",  1956,
  "021",     "Ireland",             1956,
  "006",     "Italy",               1956,
  "701",     "Japan",               1956,
  "007",     "Netherlands",         1956,
  "008",     "Norway",              1956,
  "050",     "Spain",               1956,
  "010",     "Sweden",              1956,
  "011",     "Switzerland",         1956,
  "012",     "United Kingdom",      1956,
  "302",     "United States",       1956,
  "087",     "Russian Federation",  1997,
  "546",     "Israel",              2014,
  "742",     "Korea, Republic of",  2016,
  "431",     "Brazil",              2016
)
cat("PC members:", nrow(pc_mapping), "\n")
```

---

# 7. Build the creditor-level panel

Each row = one country × year × creditor_type.

```{r denominator}
KEY <- c("country_code", "country_name", "year", "income_group")

# Denominator: total PPG + IMF credit
total_ppg <- ids_long %>%
  filter(series_code == "DT.DOD.DPPG.CD", ca_code == "WLD") %>%
  select(all_of(KEY), total_ppg = value)

imf_denom <- ids_long %>%
  filter(series_code == "DT.DOD.DIMF.CD", ca_code == "WLD") %>%
  select(all_of(KEY), imf_value = value)

denominator <- total_ppg %>%
  left_join(imf_denom, by = KEY) %>%
  mutate(imf_value  = replace_na(imf_value, 0),
         denominator = total_ppg + imf_value) %>%
  filter(denominator > 0)

cat("Denominator rows:", nrow(denominator), "\n")
```

```{r bilateral-components}
# Total bilateral (WLD aggregate)
blat_total <- ids_long %>%
  filter(series_code == "DT.DOD.BLAT.CD", ca_code == "WLD") %>%
  select(all_of(KEY), blat_total = value)

# PC bilateral: sum of creditors that were PC members at the time
blat_pc <- ids_long %>%
  filter(series_code == "DT.DOD.BLAT.CD",
         ca_code %in% pc_mapping$ca_code) %>%
  left_join(pc_mapping %>% select(ca_code, accession_year), by = "ca_code") %>%
  filter(year >= accession_year) %>%
  group_by(across(all_of(KEY))) %>%
  summarise(blat_pc = sum(value, na.rm = TRUE), .groups = "drop")

# Non-PC bilateral = total - PC
bilateral <- blat_total %>%
  left_join(blat_pc, by = KEY) %>%
  mutate(blat_pc    = replace_na(blat_pc, 0),
         blat_nonpc = pmax(blat_total - blat_pc, 0))

cat("Bilateral rows:", nrow(bilateral), "\n")
```

```{r private-components}
bonds <- ids_long %>%
  filter(series_code == "DT.DOD.PBND.CD", ca_code == "WLD") %>%
  select(all_of(KEY), value)

banks <- ids_long %>%
  filter(series_code == "DT.DOD.PCBK.CD", ca_code == "WLD") %>%
  select(all_of(KEY), value)

other_private <- ids_long %>%
  filter(series_code == "DT.DOD.PROP.CD", ca_code == "WLD") %>%
  select(all_of(KEY), value)

multilateral <- ids_long %>%
  filter(series_code == "DT.DOD.MLAT.CD", ca_code == "WLD") %>%
  select(all_of(KEY), value)

imf_comp <- ids_long %>%
  filter(series_code == "DT.DOD.DIMF.CD", ca_code == "WLD") %>%
  select(all_of(KEY), value)

cat("Bonds:", nrow(bonds), "| Banks:", nrow(banks),
    "| Other private:", nrow(other_private),
    "| Multilateral:", nrow(multilateral), "\n")
```

```{r build-long-panel}
# Stack all creditor types into one long dataframe
panel_long <- bind_rows(
  bilateral %>% select(all_of(KEY), value = blat_pc)    %>% mutate(creditor_type = "Bilateral PC"),
  bilateral %>% select(all_of(KEY), value = blat_nonpc) %>% mutate(creditor_type = "Bilateral Non-PC"),
  bonds         %>% mutate(creditor_type = "Bonds"),
  banks         %>% mutate(creditor_type = "Commercial Banks"),
  other_private %>% mutate(creditor_type = "Other Private"),
  multilateral  %>% mutate(creditor_type = "Multilateral"),
  imf_comp      %>% mutate(creditor_type = "IMF Credit")
) %>%
  filter(!is.na(value), value >= 0)

cat("Long panel rows:", format(nrow(panel_long), big.mark = ","), "\n")
```

---

# 8. Apply resolution mechanism classification with case_when

Rules:
- **Paris Club bilateral** → covered (Paris Club mechanism)
- **Non-PC bilateral** → not covered before 2020; covered (CF) from 2020
- **Commercial banks** → not covered before 1976; covered (London Club) from 1976
- **Bonds** → not covered before 2003; covered (CACs) from 2003
- **Other private** → never covered (no effective mechanism)
- **Multilateral + IMF** → never covered (outside framework)

```{r case-when}
panel_classified <- panel_long %>%
  mutate(
    mechanism = case_when(
      creditor_type == "Bilateral PC"                         ~ "Official covered (Paris Club / CF)",
      creditor_type == "Bilateral Non-PC" & year >= 2020     ~ "Official covered (Paris Club / CF)",
      creditor_type == "Bilateral Non-PC" & year <  2020     ~ "Official outside framework",
      creditor_type == "Commercial Banks" & year >= 1976     ~ "Private banks covered (London Club)",
      creditor_type == "Commercial Banks" & year <  1976     ~ "Private outside framework",
      creditor_type == "Bonds"            & year >= 2003     ~ "Bond debt covered (CACs)",
      creditor_type == "Bonds"            & year <  2003     ~ "Private outside framework",
      creditor_type == "Other Private"                        ~ "Private outside framework",
      creditor_type %in% c("Multilateral", "IMF Credit")     ~ "Multilateral / IMF outside",
      TRUE                                                    ~ "Private outside framework"
    )
  )

cat("Mechanism distribution:\n")
panel_classified %>%
  count(creditor_type, mechanism) %>%
  as.data.frame() %>%
  print()
```

---

# 9. Share of total per country × year × mechanism

```{r shares}
panel_shares <- panel_classified %>%
  left_join(denominator %>% select(all_of(KEY), denominator), by = KEY) %>%
  filter(!is.na(denominator), denominator > 0) %>%
  mutate(share = value / denominator * 100) %>%
  group_by(across(all_of(c(KEY, "mechanism")))) %>%
  summarise(share = sum(share, na.rm = TRUE), .groups = "drop")

cat("Panel shares rows:", format(nrow(panel_shares), big.mark = ","), "\n")
```

---

# 10. Annual median across LICs + renormalize to 100 %

```{r annual-lic}
mechs <- c(
  "Official covered (Paris Club / CF)",
  "Private banks covered (London Club)",
  "Bond debt covered (CACs)",
  "Official outside framework",
  "Private outside framework",
  "Multilateral / IMF outside"
)

annual_lic <- panel_shares %>%
  filter(income_group == "LIC") %>%
  group_by(year, mechanism) %>%
  summarise(median_share = median(share, na.rm = TRUE),
            n_countries  = n_distinct(country_code),
            .groups = "drop")

# Renormalize so stacked areas sum to 100 % each year
annual_lic_norm <- annual_lic %>%
  group_by(year) %>%
  mutate(row_sum     = sum(median_share, na.rm = TRUE),
         pct_norm    = median_share / row_sum * 100) %>%
  ungroup()

# Verification table
annual_lic_norm %>%
  select(year, mechanism, pct_norm) %>%
  pivot_wider(names_from = mechanism, values_from = pct_norm, values_fill = 0) %>%
  filter(year %in% c(1970, 1975, 1976, 1977, 2002, 2003, 2004,
                     2019, 2020, 2021, 2024)) %>%
  as.data.frame() %>%
  print()
```

---

# 11. Chart

```{r chart, fig.width=10, fig.height=6}
color_map <- c(
  "Official covered (Paris Club / CF)"  = "#2d6a4f",
  "Private banks covered (London Club)" = "#52b788",
  "Bond debt covered (CACs)"            = "#74c0c4",
  "Official outside framework"          = "#e07b54",
  "Private outside framework"           = "#8b4513",
  "Multilateral / IMF outside"          = "#5a4a42"
)

# Stacking: multilateral at bottom, official covered at top
stack_levels <- c(
  "Multilateral / IMF outside",
  "Private outside framework",
  "Official outside framework",
  "Bond debt covered (CACs)",
  "Private banks covered (London Club)",
  "Official covered (Paris Club / CF)"
)

plot_data <- annual_lic_norm %>%
  mutate(mechanism = factor(mechanism, levels = stack_levels))

# Transition years where jumps should appear
transitions <- data.frame(
  xintercept = c(1976, 2003, 2020),
  label      = c("London Club\n(1976)", "CACs\n(2003)", "Common\nFramework\n(2020)")
)

p7 <- ggplot(plot_data, aes(x = year, y = pct_norm, fill = mechanism)) +
  geom_area(position = "stack", alpha = 0.88) +
  geom_vline(data = transitions,
             aes(xintercept = xintercept),
             linetype = "dashed", color = "grey20", linewidth = 0.7) +
  geom_text(data = transitions,
            aes(x = xintercept + 0.3, y = 97, label = label),
            inherit.aes = FALSE,
            hjust = 0, vjust = 1, size = 2.8, color = "grey20", lineheight = 0.85) +
  scale_fill_manual(values = color_map, name = NULL,
                    guide  = guide_legend(nrow = 2, reverse = TRUE)) +
  scale_x_continuous(breaks = seq(1970, 2025, 5), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%"),
                     expand = c(0, 0)) +
  labs(
    title    = "Share of Debt Covered vs Outside Resolution Mechanisms",
    subtitle = "Median across LIC countries, normalized to 100%",
    x        = "Year",
    y        = "Percent of total external debt"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(size = 9, color = "grey40"),
    legend.text        = element_text(size = 9)
  )

p7
```

---

# 12. Save chart

```{r save-chart}
ggsave(file.path(output_dir, "ex7_resolution_mechanisms.png"),
       p7, width = 10, height = 6, dpi = 300)
cat("Chart saved\n")
```

---

# 13. Export to Excel

```{r export}
wb_out <- createWorkbook()

# Normalized chart data
addWorksheet(wb_out, "Chart - LIC (normalized)")
writeData(wb_out, "Chart - LIC (normalized)",
  annual_lic_norm %>%
    select(year, mechanism, n_countries, median_share, pct_norm) %>%
    pivot_wider(names_from  = mechanism,
                values_from = c(median_share, pct_norm),
                values_fill = 0) %>%
    arrange(year))

# Country-level detail
addWorksheet(wb_out, "Panel - LIC detail")
writeData(wb_out, "Panel - LIC detail",
  panel_shares %>%
    filter(income_group == "LIC") %>%
    arrange(country_name, year, mechanism))

# Paris Club reference
addWorksheet(wb_out, "PC Members")
writeData(wb_out, "PC Members", pc_mapping)

excel_file <- file.path(output_dir, "ex7_resolution_mechanisms.xlsx")
saveWorkbook(wb_out, excel_file, overwrite = TRUE)
cat("Excel saved:", excel_file, "\n")
```
