---
title: "Exercise 7: Debt Resolution Mechanisms"
subtitle: "BP3 — Fund's Role in International Debt Architecture"
output: html_notebook
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(openxlsx)
library(scales)
cat("Libraries loaded\n")
```

```{r paths}
ids_dir             <- "C:/Users/gsoto/International Monetary Fund (PRD)/IEO-DIL - Documents/Data/WB IDS/WB IDS25"
characteristics_dir <- "U:/dil/main-characteristics"
output_dir          <- "U:/dil/bp3/output/resolution-mechanisms"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# 1. Load IDS (6 Excel files)

```{r load-ids}
files <- c("A_D.xlsx", "E_K.xlsx", "L_M.xlsx", "N_Q.xlsx", "R_U.xlsx", "V_Z.xlsx")
ids_list <- list()
for (f in files) {
  cat("Loading", f, "...\n")
  ids_list[[f]] <- read.xlsx(file.path(ids_dir, f))
  cat("  done:", format(nrow(ids_list[[f]]), big.mark = ","), "rows\n")
}
year_cols <- names(ids_list[[1]])[grepl("^\\d{4}$", names(ids_list[[1]]))]
for (i in seq_along(ids_list)) {
  ids_list[[i]][year_cols] <- lapply(ids_list[[i]][year_cols], as.character)
}
ids_raw <- bind_rows(ids_list)
rm(ids_list); gc()
cat("\nCombined:", format(nrow(ids_raw), big.mark = ","), "rows |",
    n_distinct(ids_raw$Country.Name), "countries\n")
```

---

# 2. Filter to key series

```{r filter-series}
keep_series <- c(
  "DT.DOD.BLAT.CD",   # PPG bilateral (by counterpart country — used to split PC vs non-PC)
  "DT.DOD.MLAT.CD",   # PPG multilateral (WLD = total; always outside framework)
  "DT.DOD.PRVT.CD",   # PPG private (BND = bonds, PBNKS = commercial banks, WLD = total)
  "DT.DOD.DPPG.CD",   # Total PPG DOD (denominator base)
  "DT.DOD.DIMF.CD"    # IMF credit DOD (added to denominator; always outside framework)
)

ids_raw <- ids_raw %>% filter(Series.Code %in% keep_series)
cat("Filtered:", format(nrow(ids_raw), big.mark = ","), "rows\n")
ids_raw %>% count(Series.Code, Series.Name) %>% print(n = Inf)
```

---

# 3. Remove aggregates

```{r remove-aggregates}
remove_these <- c(
  "East Asia & Pacific (excluding high income)",
  "Europe & Central Asia (excluding high income)",
  "IDA only", "IDA total",
  "Latin America & Caribbean (excluding high income)",
  "Least developed countries: UN classification",
  "Low & middle income", "Low income", "Lower middle income",
  "Middle East & North Africa (excluding high income)",
  "Middle East, North Africa, Afghanistan & Pakistan (excluding high income)",
  "Middle income", "South Asia",
  "Sub-Saharan Africa (excluding high income)",
  "Upper middle income",
  "Heavily indebted poor countries (HIPC)", "World"
)
ids_raw <- ids_raw %>% filter(!Country.Name %in% remove_these)
cat("Countries after removing aggregates:", n_distinct(ids_raw$Country.Name), "\n")
```

---

# 4. Reshape to long format

```{r reshape}
year_cols <- names(ids_raw)[grepl("^\\d{4}$", names(ids_raw))]

ids_long <- ids_raw %>%
  pivot_longer(cols = all_of(year_cols), names_to = "year", values_to = "value") %>%
  mutate(year  = as.integer(year),
         value = as.numeric(value)) %>%
  filter(year >= 1970, year <= 2024, !is.na(value)) %>%
  rename(country_code = Country.Code,
         country_name = Country.Name,
         ca_name      = `Counterpart-Area.Name`,
         ca_code      = `Counterpart-Area.Code`,
         series_name  = Series.Name,
         series_code  = Series.Code) %>%
  mutate(ca_code      = trimws(ca_code),
         country_code = trimws(country_code))

rm(ids_raw); gc()
cat("ids_long:", format(nrow(ids_long), big.mark = ","), "rows\n")
```

---

# 5. LIC classification

```{r lic-classification}
chars   <- read.xlsx(file.path(characteristics_dir, "country-characteristics.xlsx"), sheet = "main")
lic_iso <- trimws(chars$iso)
lic_iso <- lic_iso[!is.na(lic_iso) & lic_iso != ""]

ids_long <- ids_long %>%
  mutate(income_group = if_else(country_code %in% lic_iso, "LIC", "MIC"))

cat("LIC:", n_distinct(ids_long$country_code[ids_long$income_group == "LIC"]),
    "| MIC:", n_distinct(ids_long$country_code[ids_long$income_group == "MIC"]), "\n")
```

---

# 6. Paris Club mapping

PC members and accession years — identical to ex_arrears_v3.

```{r pc-mapping}
pc_mapping <- tribble(
  ~ca_code,  ~pc_name,              ~accession_year,
  "801",     "Australia",           1956,
  "001",     "Austria",             1956,
  "002",     "Belgium",             1956,
  "301",     "Canada",              1956,
  "003",     "Denmark",             1956,
  "018",     "Finland",             1956,
  "004",     "France",              1956,
  "005",     "Germany, Fed. Rep.",  1956,
  "021",     "Ireland",             1956,
  "006",     "Italy",               1956,
  "701",     "Japan",               1956,
  "007",     "Netherlands",         1956,
  "008",     "Norway",              1956,
  "050",     "Spain",               1956,
  "010",     "Sweden",              1956,
  "011",     "Switzerland",         1956,
  "012",     "United Kingdom",      1956,
  "302",     "United States",       1956,
  "087",     "Russian Federation",  1997,
  "546",     "Israel",              2014,
  "742",     "Korea, Republic of",  2016,
  "431",     "Brazil",              2016
)
cat("PC members:", nrow(pc_mapping), "\n")
```

---

# 7. Build debt components

## 7a. Bilateral: PC vs. non-PC

PC bilateral = sum of debt to creditors that were PC members at the time of the observation.
Non-PC bilateral = total bilateral (WLD) minus PC bilateral.

```{r bilateral}
KEY <- c("country_code", "country_name", "year", "income_group")

blat_total <- ids_long %>%
  filter(series_code == "DT.DOD.BLAT.CD", ca_code == "WLD") %>%
  select(all_of(KEY), blat_total = value)

blat_pc <- ids_long %>%
  filter(series_code == "DT.DOD.BLAT.CD",
         ca_code %in% pc_mapping$ca_code) %>%
  left_join(pc_mapping %>% select(ca_code, accession_year), by = "ca_code") %>%
  filter(year >= accession_year) %>%
  group_by(across(all_of(KEY))) %>%
  summarise(blat_pc = sum(value, na.rm = TRUE), .groups = "drop")

blat <- blat_total %>%
  left_join(blat_pc, by = KEY) %>%
  mutate(blat_pc    = replace_na(blat_pc, 0),
         blat_nonpc = pmax(blat_total - blat_pc, 0))

cat("Bilateral rows:", nrow(blat), "\n")
```

## 7b. Multilateral total

```{r multilateral}
mlat <- ids_long %>%
  filter(series_code == "DT.DOD.MLAT.CD", ca_code == "WLD") %>%
  select(all_of(KEY), mlat_total = value)

cat("Multilateral rows:", nrow(mlat), "\n")
```

## 7c. Private: bonds, commercial banks, other

Bonds = `DT.DOD.PRVT.CD` with `ca_code == "BND"`.
Commercial banks = `DT.DOD.PRVT.CD` with `ca_code == "PBNKS"`.
Other private = total private (WLD) minus bonds minus banks.

```{r private}
bonds <- ids_long %>%
  filter(series_code == "DT.DOD.PRVT.CD", ca_code == "BND") %>%
  group_by(across(all_of(KEY))) %>%
  summarise(bonds_value = sum(value, na.rm = TRUE), .groups = "drop")

banks <- ids_long %>%
  filter(series_code == "DT.DOD.PRVT.CD", ca_code == "PBNKS") %>%
  group_by(across(all_of(KEY))) %>%
  summarise(banks_value = sum(value, na.rm = TRUE), .groups = "drop")

prvt_total <- ids_long %>%
  filter(series_code == "DT.DOD.PRVT.CD", ca_code == "WLD") %>%
  select(all_of(KEY), prvt_total = value)

prvt <- prvt_total %>%
  left_join(bonds, by = KEY) %>%
  left_join(banks, by = KEY) %>%
  mutate(bonds_value   = replace_na(bonds_value, 0),
         banks_value   = replace_na(banks_value, 0),
         other_private = pmax(prvt_total - bonds_value - banks_value, 0))

cat("Private rows:", nrow(prvt), "\n")
cat("  Bonds (BND) obs:   ", nrow(bonds), "\n")
cat("  Banks (PBNKS) obs: ", nrow(banks), "\n")
```

## 7d. Verify private counterparts in data

```{r verify-private-ca}
ids_long %>%
  filter(series_code == "DT.DOD.PRVT.CD") %>%
  distinct(ca_code, ca_name) %>%
  arrange(ca_code) %>%
  print(n = Inf)
```

## 7e. Total PPG and IMF credit

```{r ppg-imf}
total_ppg <- ids_long %>%
  filter(series_code == "DT.DOD.DPPG.CD", ca_code == "WLD") %>%
  select(all_of(KEY), total_ppg = value)

imf <- ids_long %>%
  filter(series_code == "DT.DOD.DIMF.CD", ca_code == "WLD") %>%
  select(all_of(KEY), imf_value = value)

cat("Total PPG rows:", nrow(total_ppg), "| IMF rows:", nrow(imf), "\n")
```

---

# 8. Merge panel

```{r merge}
panel <- total_ppg %>%
  left_join(blat,  by = KEY) %>%
  left_join(mlat,  by = KEY) %>%
  left_join(prvt,  by = KEY) %>%
  left_join(imf,   by = KEY) %>%
  mutate(across(c(blat_total, blat_pc, blat_nonpc,
                  mlat_total,
                  prvt_total, bonds_value, banks_value, other_private,
                  imf_value),
                ~ replace_na(., 0))) %>%
  mutate(denominator = total_ppg + imf_value) %>%
  filter(denominator > 0)

cat("Panel:", format(nrow(panel), big.mark = ","), "rows |",
    n_distinct(panel$country_code), "countries |",
    min(panel$year), "–", max(panel$year), "\n")
```

---

# 9. Resolution mechanism mapping

Rules (from IEO team):

| Category | Rule |
|---|---|
| Official covered (Paris Club / CF) | PC bilateral before 2020; **all** bilateral from 2020 (Common Framework) |
| Official outside framework | Non-PC bilateral before 2020; **0** from 2020 |
| Private banks covered (London Club) | Commercial banks (PBNKS) from 1976; 0 before |
| Bond debt covered (CACs) | Bonds (BND) from 2003; 0 before |
| Multilateral / IMF outside | All multilateral (PCS + non-PCS) + IMF credit; always outside |
| Private outside framework | Other private (always) + banks before 1976 + bonds before 2003 |

```{r mapping}
panel <- panel %>%
  mutate(
    official_covered = if_else(year >= 2020, blat_total, blat_pc),
    official_outside = if_else(year >= 2020, 0,          blat_nonpc),
    banks_covered    = if_else(year >= 1976, banks_value, 0),
    bonds_covered    = if_else(year >= 2003, bonds_value, 0),
    multilateral_out = mlat_total + imf_value,
    private_outside  = other_private +
                       if_else(year < 1976, banks_value, 0) +
                       if_else(year < 2003, bonds_value, 0)
  )

cats_raw <- c("official_covered", "banks_covered", "bonds_covered",
              "official_outside", "private_outside", "multilateral_out")
```

---

# 10. Convert to % of denominator

```{r pct}
cats_pct <- paste0("pct_", cats_raw)

panel_pct <- panel %>%
  mutate(across(all_of(cats_raw),
                ~ . / denominator * 100,
                .names = "pct_{.col}"))
```

---

# 11. Quick diagnostic — LIC coverage check

```{r diagnostic}
panel_pct %>%
  filter(income_group == "LIC") %>%
  group_by(year) %>%
  summarise(across(all_of(cats_pct), ~ median(., na.rm = TRUE)),
            n_countries = n_distinct(country_code),
            .groups = "drop") %>%
  mutate(row_sum = rowSums(across(all_of(cats_pct)))) %>%
  filter(year %in% c(1970, 1976, 1980, 1990, 2000, 2003, 2010, 2015, 2020, 2024)) %>%
  print(width = Inf)
```

---

# 12. Annual median for LICs + renormalize to 100 %

The median of component percentages across countries does not sum to 100 %. We renormalize each year so the stacked areas fill the chart correctly.

```{r annual-lic}
annual_lic <- panel_pct %>%
  filter(income_group == "LIC") %>%
  group_by(year) %>%
  summarise(across(all_of(cats_pct), ~ median(., na.rm = TRUE)),
            n_countries = n_distinct(country_code),
            .groups = "drop")

annual_lic_norm <- annual_lic %>%
  mutate(row_sum = rowSums(across(all_of(cats_pct))),
         across(all_of(cats_pct), ~ . / row_sum * 100)) %>%
  select(-row_sum)

print(annual_lic_norm, n = Inf)
```

---

# 13. Chart

Stacking order: multilateral at bottom, official covered at top.
Vertical dashed lines mark the three key framework dates.

```{r chart, fig.width=10, fig.height=6}
label_map <- c(
  pct_official_covered = "Official covered (Paris Club / CF)",
  pct_banks_covered    = "Private banks covered (London Club)",
  pct_bonds_covered    = "Bond debt covered (CACs)",
  pct_official_outside = "Official outside framework",
  pct_private_outside  = "Private outside framework",
  pct_multilateral_out = "Multilateral / IMF outside"
)

color_map <- c(
  "Official covered (Paris Club / CF)"  = "#2d6a4f",
  "Private banks covered (London Club)" = "#52b788",
  "Bond debt covered (CACs)"            = "#74c0c4",
  "Official outside framework"          = "#e07b54",
  "Private outside framework"           = "#8b4513",
  "Multilateral / IMF outside"          = "#5a4a42"
)

# Factor levels: first = bottom of stack, last = top of stack
stack_levels <- c(
  "Multilateral / IMF outside",
  "Private outside framework",
  "Official outside framework",
  "Bond debt covered (CACs)",
  "Private banks covered (London Club)",
  "Official covered (Paris Club / CF)"
)

plot_data <- annual_lic_norm %>%
  pivot_longer(cols = all_of(cats_pct), names_to = "mechanism", values_to = "pct") %>%
  mutate(mechanism = recode(mechanism, !!!label_map),
         mechanism = factor(mechanism, levels = stack_levels))

p7 <- ggplot(plot_data, aes(x = year, y = pct, fill = mechanism)) +
  geom_area(position = "stack", alpha = 0.88) +
  geom_vline(xintercept = c(1976, 2003, 2020),
             linetype = "dashed", color = "white", linewidth = 0.7) +
  scale_fill_manual(values = color_map, name = NULL,
                    guide  = guide_legend(nrow = 2, reverse = TRUE)) +
  scale_x_continuous(breaks = seq(1970, 2025, 5), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%"),
                     expand = c(0, 0)) +
  labs(
    title    = "Share of Debt Covered vs Outside Resolution Mechanisms",
    subtitle = paste0("Median across LIC countries, normalized to 100%\n",
                      "Dashed lines: 1976 (London Club), 2003 (CACs), 2020 (Common Framework)"),
    x        = "Year",
    y        = "Percent of total external debt"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(size = 9, color = "grey40"),
    legend.text        = element_text(size = 9)
  )

p7
```

---

# 14. Save chart

```{r save-chart}
ggsave(file.path(output_dir, "ex7_resolution_mechanisms.png"),
       p7, width = 10, height = 6, dpi = 300)
cat("Chart saved\n")
```

---

# 15. Export to Excel

```{r export}
wb_out <- createWorkbook()

# Chart-ready data (normalized, for replication)
addWorksheet(wb_out, "Chart - LIC (normalized)")
writeData(wb_out, "Chart - LIC (normalized)",
  annual_lic_norm %>%
    select(year, n_countries, all_of(cats_pct)) %>%
    rename(
      `Official covered (Paris Club / CF)`  = pct_official_covered,
      `Private banks covered (London Club)` = pct_banks_covered,
      `Bond debt covered (CACs)`            = pct_bonds_covered,
      `Official outside framework`          = pct_official_outside,
      `Private outside framework`           = pct_private_outside,
      `Multilateral / IMF outside`          = pct_multilateral_out
    ))

# Raw medians (before normalization)
addWorksheet(wb_out, "Chart - LIC (raw median %)")
writeData(wb_out, "Chart - LIC (raw median %)",
  annual_lic %>%
    select(year, n_countries, all_of(cats_pct)) %>%
    rename(
      `Official covered (Paris Club / CF)`  = pct_official_covered,
      `Private banks covered (London Club)` = pct_banks_covered,
      `Bond debt covered (CACs)`            = pct_bonds_covered,
      `Official outside framework`          = pct_official_outside,
      `Private outside framework`           = pct_private_outside,
      `Multilateral / IMF outside`          = pct_multilateral_out
    ))

# Country-level detail for LICs
addWorksheet(wb_out, "Panel - LIC detail")
writeData(wb_out, "Panel - LIC detail",
  panel_pct %>%
    filter(income_group == "LIC") %>%
    select(country_name, country_code, year,
           blat_total, blat_pc, blat_nonpc,
           mlat_total, bonds_value, banks_value, other_private, imf_value,
           denominator,
           all_of(cats_raw),
           all_of(cats_pct)) %>%
    arrange(country_name, year))

# Reference: Paris Club member list
addWorksheet(wb_out, "PC Members")
writeData(wb_out, "PC Members", pc_mapping)

excel_file <- file.path(output_dir, "ex7_resolution_mechanisms.xlsx")
saveWorkbook(wb_out, excel_file, overwrite = TRUE)
cat("Excel saved:", excel_file, "\n")
```
